Supporting Information: Thermal sensitivity across forest vertical profiles: patterns, mechanisms, and ecological implications

### 2021-10-01 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

Light, specifically the proportion of incident light and photosynthetically active radiation (PAR), decreases from the canopy top to the forest floor, with the profile shape modified by leaf area density, canopy height, and canopy structure across species and forest types [Fig. `r fig_NEON_vertical`a-d, Supporting Information Figure `r SIfig_NEON_all`, @koike_leaf_2001].
This gradient is more pronounced in dense canopies, including broad-leaf and mixed forests (e.g., SCBI, SERC and HARV, Fig. `r fig_NEON_vertical`d), than in open canopy forests, including many conifer forests [e.g., OSBS, PUUM, and WREF, Fig. `r fig_NEON_vertical`d, Supporting Information Figure `r SIfig_NEON_all`, @chazdonPhotosyntheticLightEnvironments1984; @aussenacInteractionsForestStands2000a; @smithSeasonalDroughtrelatedChanges2019; @tymenQuantifyingMicroenvironmentalVariation2017;  @parkerConsequencesEnvironmentalHeterogeneity2019; @bartemucciOverstoryInfluencesLight2006; @baldocchiSeasonalVariationEnergy1997]. 
Variability in the light environment decreases with height, with the shaded understory being highly influenced by sunflecks, or brief increases in solar radiation caused by small canopy gaps and wind-induced canopy movements [@waySunflecksTreesForests2012]. 
Likewise, the mid-canopy experiences a highly variable light environment due to the sun's passage across a dynamically structured canopy surface [**REF- Ty or Marielle**]
The implication of this light gradient is that upper canopy leaves have higher potential photosynthetic rate (see section 4.2), but also greater propensity towards elevated $T_{leaf}$ (see section 2.2).

@@

Relative humidity (RH) tends to be higher in the understory and decreases with height, although this trend is absent in open forests [Fig. `r fig_NEON_vertical`f, Supporting Information Figure `r SIfig_NEON_all`, e.g., @juckerCanopyStructureTopography2018; @mcgregorTreeHeightLeaf2021].
Dense-canopy forests maintain higher daily maximum RH in the understory than open forests and nearby open areas, an effect that is greater in wetter conditions, and warmer months [@vonarxSpatiotemporalEffectsForest2012a; @hanberryOpenForestEcosystems2020a]. 
In combination with $T_{air}$, RH determines vapor pressure deficit (VPD)-- the driving force of water loss from leaves--which tends to be greater in the upper canopy and canopy gaps than in the understory [@fauset_differences_2018; @tymenQuantifyingMicroenvironmentalVariation2017; @niinemetsPhotosyntheticAcclimationSimultaneous2004b]. Especially when canopy height and LAI are large, maximum $T_{air}$ and VPD are reduced in the understory [@juckerCanopyStructureTopography2018].
The implication is that canopy leaves face higher $g_b$ and associated potential for water and heat loss (see section 2.2).

@@

The basic biophysical principles outlined above shape the tendency for $T_{leaf}$ to be elevated above $T_{air}$ across vertical gradients in closed-canopy forests.
High radiation at the top of the vertical profile (Fig. `r fig_NEON_vertical`) implies that canopy leaves have much greater tendency for high $T_{leaf}-T_{air}$, such that they would be expected to be warmer than understory leaves under most conditions (Fig. `r fig_leaf_T`).
However, higher wind speeds (Fig. `r fig_NEON_vertical`) reduce $T_{leaf}-T_{air}$ [Fig. `r fig_leaf_T`b, @bonan_ecological_2016; @niinemetsPhotosyntheticAcclimationSimultaneous2004b]. 
In addition, adaptive leaf traits that increase $g_b$ and $g_s$ (see "Leaf Traits" and "Leaf gas exchange and its thermal sensitivity" sections) mediate the direct effects of solar radiation on upper canopy leaves and result in higher rates of heat loss (Fig. `r fig_leaf_T`b-e). 
In contrast, greater RH and lower wind speeds in the understory may limit $\lambda E$ (through reduced $g_s$ and $g_b$), thereby increasing $T_{leaf}$ [Fig. `r fig_leaf_T`, @perezIncreasingHumidityThreatens2018; @tibbittsHumidityPlants1979b].

The propensity for $T_{leaf}$ to be elevated above $T_{air}$ is not expected to be uniform across time and space, but rather to vary with micrometeorological conditions. Under drought conditions (hot and dry, with higher-than-average solar radiation), when $g_s$ is limiting, there will be a greater propensity for $T_{leaf} - T_{air}$ to be elevated in the canopy compared to the understory (Fig.`r fig_leaf_T`).
This is because sun leaves are exposed to higher irradiance and VPD, and therefore have steeper $g_s$ limitation than shade leaves [Fig. `r fig_leaf_T`e, @leighInfluenceLeafSize2017; @fauset_differences_2018]. 
In contrast, under hot and wet conditions, higher wind speeds and $g_b$ enable cooling in the upper canopy, whereas lower wind and $g_b$ in the understory allow much less evaporative cooling  [Fig. `r fig_leaf_T`b,d, @robertsStomatalBoundaryLayerConductances1990a; @martin_boundary_1999; @leighInfluenceLeafSize2017;  @songTreeSurfaceTemperature2020]. 
Thus, while understory and within-canopy shade leaves can remain cooler under lower radiation, their environment is not conducive to shedding excess heat, which in combination with their physiological propensities may result in above-optimal $T_{leaf}$ under slight heat or drought stress, or when exposed to higher levels of radiation, for example during sunflecks [@leighInfluenceLeafSize2017; @songTreeSurfaceTemperature2020; @schymanskiStomatalControlLeaf2013b; @zweifelMiddayStomatalClosure2002].
In addition to lower wind speeds (Fig. `r fig_NEON_vertical`), higher RH in the understory and inner canopy (Fig. `r fig_NEON_vertical`) would also inhibit cooling, as $\lambda E$ is stifled under high RH, resulting in  greater $T_{leaf}$-$T_{air}$ [Fig. `r fig_leaf_T`c, @tibbittsHumidityPlants1979b; @songTreeSurfaceTemperature2020;  @perezIncreasingHumidityThreatens2018; @konradLeafTemperatureIts2021; @dietzVerticalPatternsDuration2007].

@@

Most leaf biochemical traits also vary across light and height gradients (Table `r table_leaf_traits`), both shaping and resulting from gradients in metabolism (see "Leaf gas exchange and its thermal sensitivity" section, Table `r table_leaf_metabolism`).
Sun leaves have higher concentrations (per unit area) of elements such as nitrogen ($N_{area}$) and phosphorus ($P_{area}$) that are critical to leaf metabolism, including respiration [@meirLeafRespirationTwo2001;@weerasingheCanopyPositionAffects2014] and photosynthetic processes [@scartazzaInvestigatingEuropeanBeech2016; @niinemetsWithincanopyVariationRate2004a; @weerasingheCanopyPositionAffects2014].
Increases in chlorophyll a/b ratios with height reflect the greater light availability in the upper canopy, while greater chlorophyll concentrations at lower heights within the canopy give shade leaves greater PAR absorptance efficiency (Table `r table_leaf_traits`).
Higher photosynthetic rates (per unit leaf area) and more frequent stomatal closure higher (see "Leaf gas exchange and its thermal sensitivity" section, Table `r table_leaf_metabolism`) result in lower intercellular CO~2~ concentrations and higher $\delta^{13}$C concentrations in leaf tissues (Table `r table_leaf_traits`).

Biochemical protection against foliage light and heat damage also increases with irradiance, and thus tends to be higher in the upper canopy than in the understory (Table `r table_leaf_traits`). 
Carotenoids (e.g., beta carotene and lutein), including xanthophyll cycle pigments (i.e., violaxanthin, antheraxanthin and zeaxanthin, VAZ), can play a role in antioxidant scavenging and converting excess excitation energy into heat to reduce photoinhibition [@niinemetsPhotosynthesisResourceDistribution2007a; @mathurPhotosyntheticEfficiencySun2018a; @niinemetsAcclimationHighIrradiance1998]. 
The dissipation of excess light energy is also important for leaf acclimation to higher temperatures if high $T_{leaf}$ impairs the photochemical energy dissipation pathway [@havauxTemperaturedependentAdjustmentThermal1996]. 
Capacity for heat-sensitive emissions of volatile organic compounds (VOCs), including isoprenes and monoterpenes, enhances photosynthetic thermal tolerance by regulating antioxidant defenses and other metabolic processes [@sharkeyIsopreneEmissionPlants2008; @taylorCapacityEmitIsoprene2019; @monsonLeafIsopreneEmission2021; **REF ON MONOTERPENES?**]. 
VOC emission capacity is a species-level trait [**REFS**], such that intrapspecific variation across across vertical gradients are not applicable, but frequency of this trait varies across the vertical gradient (see section 3.2), as do emission rates [see section 4.4, @taylorNewFieldInstrument2021].

>>>>

Light, specifically the proportion of incident light and photosynthetically active radiation (PAR), decreases from the canopy top to the forest floor, with the profile shape modified by leaf area density, canopy height, and canopy structure across species and forest types [Fig. `r fig_NEON_vertical`a-d, Supporting Information Figure `r SIfig_NEON_all`, @koike_leaf_2001].
This gradient is more pronounced in dense canopies, including broad-leaf and mixed forests (e.g., SCBI, SERC and HARV, Fig. `r fig_NEON_vertical`d), than in open canopy forests, including many conifer forests [e.g., OSBS, PUUM, and WREF, Fig. `r fig_NEON_vertical`d, Supporting Information Figure `r SIfig_NEON_all`) [@chazdonPhotosyntheticLightEnvironments1984; @aussenacInteractionsForestStands2000a; @smithSeasonalDroughtrelatedChanges2019; @tymenQuantifyingMicroenvironmentalVariation2017;  @parkerConsequencesEnvironmentalHeterogeneity2019; @bartemucciOverstoryInfluencesLight2006; @baldocchiSeasonalVariationEnergy1997]. 
Variability in the light environment decreases with height, with the shaded understory being highly influenced by sunflecks, or brief increases in solar radiation caused by small canopy gaps and wind-induced canopy movements [@waySunflecksTreesForests2012]. 
Likewise, the mid-canopy experiences a highly variable light environment due to the sun's passage across a dynamically structured canopy surface [@waySunflecksTreesForests2012].
The implication of this light gradient is that upper canopy leaves have higher potential photosynthetic rate (see section 4.2), but also greater propensity towards elevated $T_{leaf}$ (see section 2.2).

@@

Relative humidity (RH) tends to be higher in the understory and decreases with height, although this trend is absent in open forests [Fig. `r fig_NEON_vertical`f, Supporting Information Figure `r SIfig_NEON_all`, e.g., @juckerCanopyStructureTopography2018; @mcgregorTreeHeightLeaf2021].
Dense-canopy forests maintain higher daily maximum RH in the understory than open forests and nearby open areas, an effect that is greater in wetter conditions, and warmer months [@vonarxSpatiotemporalEffectsForest2012a; @hanberryOpenForestEcosystems2020a]. 
In combination with $T_{air}$, RH determines vapor pressure deficit (VPD)-- the driving force of water loss from leaves--which tends to be greater in the upper canopy and canopy gaps than in the understory [@fauset_differences_2018; @tymenQuantifyingMicroenvironmentalVariation2017; @niinemetsPhotosyntheticAcclimationSimultaneous2004]. Especially when canopy height and LAI are large, maximum $T_{air}$ and VPD are reduced in the understory [@juckerCanopyStructureTopography2018].
The implication is that canopy leaves face higher $g_b$ and associated potential for water and heat loss (see section 2.2).

@@

The basic biophysical principles outlined above shape the tendency for $T_{leaf}$ to be elevated above $T_{air}$ across vertical gradients in closed-canopy forests.
High radiation at the top of the vertical profile (Fig. `r fig_NEON_vertical`) implies that canopy leaves have much greater tendency for high $T_{leaf}-T_{air}$, such that they would be expected to be warmer than understory leaves under most conditions (Fig. `r fig_leaf_T`).
However, higher wind speeds (Fig. `r fig_NEON_vertical`) reduce $T_{leaf}-T_{air}$ [Fig. `r fig_leaf_T`b, @bonan_ecological_2016; @niinemetsPhotosyntheticAcclimationSimultaneous2004]. 
In addition, adaptive leaf traits that increase $g_b$ and $g_s$ (see "Leaf Traits" and "Leaf gas exchange and its thermal sensitivity" sections) mediate the direct effects of solar radiation on upper canopy leaves and result in higher rates of heat loss (Fig. `r fig_leaf_T`b-e). 
In contrast, greater RH and lower wind speeds in the understory may limit $\lambda E$ (through reduced $g_s$ and $g_b$), thereby increasing $T_{leaf}$ [Fig. `r fig_leaf_T`, @perezIncreasingHumidityThreatens2018; @tibbittsHumidityPlants1979b].

The propensity for $T_{leaf}$ to be elevated above $T_{air}$ is not expected to be uniform across time and space, but rather to vary with micrometeorological conditions. Under drought conditions (hot and dry, with higher-than-average solar radiation), when $g_s$ is limiting, there will be a greater propensity for $T_{leaf} - T_{air}$ to be elevated in the canopy compared to the understory (Fig.`r fig_leaf_T`).
This is because sun leaves are exposed to higher irradiance and VPD, and therefore have steeper $g_s$ limitation than shade leaves [Fig. `r fig_leaf_T`e, @leighInfluenceLeafSize2017; @fauset_differences_2018]. 
In contrast, under hot and wet conditions, higher wind speeds and $g_b$ enable cooling in the upper canopy, whereas lower wind and $g_b$ in the understory allow much less evaporative cooling  [Fig. `r fig_leaf_T`b,d, @robertsStomatalBoundaryLayerConductances1990a; @martin_boundary_1999; @leighInfluenceLeafSize2017;  @songTreeSurfaceTemperature2020]. 
Thus, while understory and within-canopy shade leaves can remain cooler under lower radiation, their environment is not conducive to shedding excess heat, which in combination with their physiological propensities may result in above-optimal $T_{leaf}$ under slight heat or drought stress, or when exposed to higher levels of radiation, for example during sunflecks [@leighInfluenceLeafSize2017; @songTreeSurfaceTemperature2020; @schymanskiStomatalControlLeaf2013b].
In addition to lower wind speeds (Fig. `r fig_NEON_vertical`), higher RH in the understory and inner canopy (Fig. `r fig_NEON_vertical`) would also inhibit cooling, as $\lambda E$ is stifled under high RH, resulting in  greater $T_{leaf}$-$T_{air}$ [Fig. `r fig_leaf_T`c, @tibbittsHumidityPlants1979b; @songTreeSurfaceTemperature2020;  @perezIncreasingHumidityThreatens2018; @konradLeafTemperatureIts2021; @dietzVerticalPatternsDuration2007].

@@

Most leaf biochemical traits also vary across light and height gradients (Table `r table_leaf_traits`), both shaping and resulting from gradients in metabolism (see "Leaf gas exchange and its thermal sensitivity" section, Table `r table_leaf_metabolism`).
Sun leaves have higher concentrations (per unit area) of elements such as nitrogen ($N_{area}$) and phosphorus ($P_{area}$) that are critical to leaf metabolism, including respiration [@meirLeafRespirationTwo2001;@weerasingheCanopyPositionAffects2014] and photosynthetic processes [@scartazzaInvestigatingEuropeanBeech2016; @niinemetsPhotosyntheticAcclimationSimultaneous2004; @weerasingheCanopyPositionAffects2014].
Increases in chlorophyll a/b ratios with height reflect the greater light availability in the upper canopy, while greater chlorophyll concentrations at lower heights within the canopy give shade leaves greater PAR absorptance efficiency (Table `r table_leaf_traits`).
Higher photosynthetic rates (per unit leaf area) and more frequent stomatal closure higher (see "Leaf gas exchange and its thermal sensitivity" section, Table `r table_leaf_metabolism`) result in lower intercellular CO~2~ concentrations and higher $\delta^{13}$C concentrations in leaf tissues (Table `r table_leaf_traits`).

Biochemical protection against foliage light and heat damage also increases with irradiance, and thus tends to be higher in the upper canopy than in the understory (Table `r table_leaf_traits`). 
Carotenoids (e.g., beta carotene and lutein), including xanthophyll cycle pigments (i.e., violaxanthin, antheraxanthin and zeaxanthin, VAZ), can play a role in antioxidant scavenging and converting excess excitation energy into heat to reduce photoinhibition [@niinemetsPhotosynthesisResourceDistribution2007a; @mathurPhotosyntheticEfficiencySun2018a; @niinemetsAcclimationHighIrradiance1998]. 
The dissipation of excess light energy is also important for leaf acclimation to higher temperatures if high $T_{leaf}$ impairs the photochemical energy dissipation pathway [@havauxTemperaturedependentAdjustmentThermal1996].

Capacity for heat-sensitive, light-dependent (photosynthetically linked) emissions of volatile organic compounds (VOCs), including isoprene and monoterpenes, enhances photosynthetic thermal tolerance by regulating antioxidant defenses and other metabolic processes [@sharkeyIsopreneEmissionPlants2008; @taylorCapacityEmitIsoprene2019; @monsonLeafIsopreneEmission2021; **REF ON MONOTERPENES?**]. 
Less than half of tree species express significant light-dependent emissions (REFS; Taylor et al., 2018), such that variation in emissions across the vertical profile can be influenced both by vertical variation in species compositions (see section 3.2) and plasticity in emission rate capacities [see section 4.4, @taylorNewFieldInstrument2021].
